Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-07-22 supplement.tex
Update supplement.tex

In our preliminary models, the unconstrained inferred split times between East
African populations and Europeans was inferred to have occurred around $40$ka,
which is more recent than we may have expected given the well-documented
presence of modern humans outside of Africa at that time \citep{}, as well as
previous genetics-based dating of splits between African and Eurasian
populations \citep{}. The unconstrained inferred split between East and West
African populations occurred about $10$ka earlier
These recent split dates seems to be at odds with archaelogical evidence for earlier
 Out-of-Africa (OOA) expansion, and with the estimated date for Neanderthal admixture
 into OOA populations.  The parameter uncertainty due to finite genome
 do not account for this difference:  in the high-likelihood models, the $95\%$ confidence 
 intervals for the east African-European split was roughly $38 \pm 5$ka, 
 and for the East/West African split was $45 \pm 10$ka, with the uncertainties 
 reflecting 2 standard deviations according to bootstrap of genomic regions.  
Thus we expect these discrepancies to result for model misspecification in the 
details of the divergence between the ancestors of OOA and east African 
populations. Our models supposes a
single split prior to the OOA event followed by constant symmetric migrations 
between eastern and OOA populations.
which is certainly an oversimplification \citep{}.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-relate-iicr-data-full-and-subset}
    \caption{
        \textbf{Inverse instantaneous coalescence rates inferred from the data,
        subset to focal populations in the model inferences.}
        Fig.~\ref{fig:supp-iicr-data} shows IICR curves from reconstructed
        genealogies inferred using a larger dataset (including nearly 1,200
        individuals) than the subset of populations we primarily focused on
        here. To assess the robustness of Relate-inferred genealogies and for a
        direct comparison to our inferred demographic models, we ran Relate on
        the 290 genomes from the Nama, Gumuz, Mende, Amhara/Oromo, and British.
        While the reconstructed IICR curves are qualitatively similar to those
        from the full dataset for these populations in the distant past, there
        are noticeable discrepancies over the recent history in the last tens of
        thousands of years. Left: Relate run on full dataset, showing focal
        populations. Right: Relate run directly on the subset dataset of focal
        populations.
    }
    \label{fig:supp-iicr-data-subset}
\end{figure}

>>>>

In our preliminary models, the unconstrained inferred split times between East
African populations and Europeans was inferred to have occurred around $40$ka,
which is more recent than we may have expected given the well-documented
presence of modern humans outside of Africa at that time \citep{}, as well as
previous genetics-based dating of splits between African and Eurasian
populations \citep{}. The unconstrained inferred split between East and West
African populations occurred about $10$ka earlier
These recent split dates seems to be at odds with archaelogical evidence for earlier
 Out-of-Africa (OOA) expansion, and with the estimated date for Neanderthal admixture
 into OOA populations. The parameter uncertainty due to finite genomes
 does not account for this difference:  in the high-likelihood models, the $95\%$ confidence 
 intervals for the east African-European split was roughly $38 \pm 5$ka, 
 and for the East/West African split was $45 \pm 10$ka, with the uncertainties 
 reflecting 2 standard deviations according to bootstrap of genomic regions.  
Thus we expect these discrepancies to result for model misspecification in the 
details of the divergence between the ancestors of OOA and east African 
populations. Our models supposes a
single split prior to the OOA event followed by constant symmetric migrations 
between eastern and OOA populations.
which is certainly an oversimplification \citep{}.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{figures/supp-relate-iicr-data-full-and-subset}
    \caption{
        \textbf{Inverse instantaneous coalescence rates inferred from the data,
        subset to focal populations in the model inferences.}
        Fig.~\ref{fig:supp-iicr-data} shows IICR curves from reconstructed
        genealogies inferred using a larger dataset (including nearly 1,200
        individuals) than the subset of populations we primarily focused on
        here. To assess the robustness of Relate-inferred genealogies and for a
        direct comparison to our inferred demographic models, we ran Relate on
        the 290 genomes from the Nama, Gumuz, Mende, Amhara/Oromo, and British individuals.
        While the reconstructed IICR curves are qualitatively similar to those
        from the full dataset for these populations in the distant past, there
        are noticeable discrepancies over the recent history in the last tens of
        thousands of years. Left: Relate run on full dataset, showing focal
        populations. Right: Relate run directly on the subset dataset of focal
        populations.
    }
    \label{fig:supp-iicr-data-subset}
\end{figure}
